---
title: "Introduction to `Rcppautodiff`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to `Rcppautodiff`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction
`Rcppautodiff` is an interface to the excellent [autodiff](https://autodiff.github.io/) modern, `C++17` header-only library for automatic differentiation. It provides an easy and intuitive way to compute derivatives automatically.

The library also comes with excellent documentation and examples. The original `C++` examples can be seen [here](https://autodiff.github.io/tutorials/). The examples in the tutorial page of [autodiff](https://autodiff.github.io/) are recreated here with `Rcppautodiff`, interspersed with comments on how to convert from `autodiff` data-types to `Rcpp` or `R` data types and back.

## Data types in `autodiff`
`autodiff` extensively uses the [Eigen](https://eigen.tuxfamily.org/index.php?title=Main_Page) library and the data types defined therein, e.g., `VectorXd` and `MatrixXd`. Additionally, is has `dual` data type (representing [dual numbers](https://en.wikipedia.org/wiki/Dual_number)) and `VectorXdual` representing a vector of dual numbers. 

### `dual` to `double` conversion
`dual` number can be converted to `double` as follows:
```{Rcpp, eval = FALSE}
dual x = 5;
double y = x.val;
```
### `VectorXreal` (or `VectorXdual`) to `Eigen::VectorXd`
`autodiff`'s data type `VectorXreal` can be converted to `Eigen::VectorXd` as follows:

```{Rcpp, eval = FALSE}
VectorXreal v(3);
v << 2.1, 4.3, 6.7;
Eigen::VectorXd y = v.cast<double>();

VectorXdual v(3);
v << 2.1, 4.3, 6.7;
Eigen::VectorXd y = v.cast<double>();
```

### `MatrixXreal` (or `MatrixXdual`) to `Eigen::MatrixXd`
Similar to the conversion for vectors (above), `autodiff`'s `MatrixXreal` can be converted to `Eigen::MatxixXd` as follows:

```{Rcpp, eval = FALSE}
MatrixXreal m(2, 2);
m << 2.1, 4.3, 6.7, 8.2;
Eigen::MatrixXd m1 = m.cast<double>();

MatrixXdual m(2, 2);
m << 2.1, 4.3, 6.7, 8.2;
Eigen::MatrixXd m1 = m.cast<double>();
```

## Example - Derivative of single-variable function

Differentiating the function
$$
f(x) = 1 + x + x*x + 1/x + log(x)
$$
w.r.t. $x$.

First load the package in an `R` session

```{r setup}
library(Rcppautodiff)
```

The following can be pasted in a separate file with `.cpp` extension. Then, use the `sourceCpp` function of `Rcpp` to compile the code and show result with input value of $x = 2.0$. 

```{Rcpp, eval = FALSE}
#include <Rcppautodiff.h>

// autodiff include
#include <autodiff/forward/dual.hpp>

using namespace Rcpp;
using namespace autodiff;

// [[Rcpp::plugins(cpp17)]]

//[[Rcpp::depends(Rcppautodiff)]]

// The single-variable function for which derivatives are needed
dual f(dual x)
{
    dual y = 1/x;     // other variables should also be dual
	return 1 + x + x*x + y + log(x);
}

// [[Rcpp::export]]
int myfun(double input){

	dual x = input;                                 // the input variable x
	dual u = f(x);                                  // the output variable u

	double dudx = derivative(f, wrt(x), at(x));     // evaluate the derivative du/dx

	Rcpp::Rcout << "u = " << u << std::endl;        // print the evaluated output u
	Rcpp::Rcout << "du/dx = " << dudx << std::endl; // print the evaluated derivative du/dx
	return(0);
}

/*** R
myfun(2.0)
*/

```


## Derivatives of a multi-variable function

Differentiating the function
$$
f(x,y,z) = 1 + x + y + z + x\cdot y + y\cdot z + x\cdot z + x\cdot y\cdot z + exp(x/y + y/z)
$$
w.r.t. $x, y, z$.

```{Rcpp, eval = FALSE}
#include <Rcppautodiff.h>

// autodiff include
#include <autodiff/forward/dual.hpp>

using namespace Rcpp;
using namespace autodiff;

// [[Rcpp::plugins(cpp17)]]

//[[Rcpp::depends(Rcppautodiff)]]

// The multi-variable function for which derivatives are needed
dual f1(dual x, dual y, dual z)
{
    return 1 + x + y + z + x*y + y*z + x*z + x*y*z + exp(x/y + y/z);
}

// [[Rcpp::export]]
int myfun1(double xin, double yin, double zin)
{
    dual x = xin;
    dual y = yin;
    dual z = zin;

    dual u = f1(x, y, z);

    double dudx = derivative(f1, wrt(x), at(x, y, z));
    double dudy = derivative(f1, wrt(y), at(x, y, z));
    double dudz = derivative(f1, wrt(z), at(x, y, z));

    std::cout << "u = " << u << std::endl;         // print the evaluated output u = f(x, y, z)
    std::cout << "du/dx = " << dudx << std::endl;  // print the evaluated derivative du/dx
    std::cout << "du/dy = " << dudy << std::endl;  // print the evaluated derivative du/dy
    std::cout << "du/dz = " << dudz << std::endl;  // print the evaluated derivative du/dz
    
    return(0);
}

/*** R
myfun1(1.0, 2.0, 3.0)
*/
```


## Derivatives of a multi-variable function with parameters

Differentiating the function
$$
f(x) = a \cdot sin(x) + b \cdot cos(x) + c \cdot sin(x) \cdot cos(x);
$$
w.r.t. $x, a, b, c$ where $a, b,$ and $c$ are constant parameters.

```{Rcpp, eval = FALSE}
#include <Rcppautodiff.h>

// autodiff include
#include <autodiff/forward/dual.hpp>

using namespace Rcpp;
using namespace autodiff;

// [[Rcpp::plugins(cpp17)]]

//[[Rcpp::depends(Rcppautodiff)]]

// A type defining parameters for a function of interest
struct Params
{
    dual a;
    dual b;
    dual c;
};

// The function that depends on parameters for which derivatives are needed
dual f2(dual x, const Params& params)
{
    return params.a * sin(x) + params.b * cos(x) + params.c * sin(x)*cos(x);
}

// [[Rcpp::export]]
int myfun2(double a_input, double b_input, double c_input, double x_input)
{
    Params params;   // initialize the parameter variables
    
    dual a = a_input;
    params.a = a;    // the parameter a of type dual, not double!
    
    dual b = b_input;
    params.b = b;  // the parameter b of type dual, not double!
    
    dual c = c_input;
    params.c = c;  // the parameter c of type dual, not double!

    dual x = x_input;  // the input variable x

    dual u = f2(x, params);  // the output variable u

    double dudx = derivative(f2, wrt(x), at(x, params));        // evaluate the derivative du/dx
    double duda = derivative(f2, wrt(params.a), at(x, params)); // evaluate the derivative du/da
    double dudb = derivative(f2, wrt(params.b), at(x, params)); // evaluate the derivative du/db
    double dudc = derivative(f2, wrt(params.c), at(x, params)); // evaluate the derivative du/dc

    std::cout << "u = " << u << std::endl;         // print the evaluated output u
    std::cout << "du/dx = " << dudx << std::endl;  // print the evaluated derivative du/dx
    std::cout << "du/da = " << duda << std::endl;  // print the evaluated derivative du/da
    std::cout << "du/db = " << dudb << std::endl;  // print the evaluated derivative du/db
    std::cout << "du/dc = " << dudc << std::endl;  // print the evaluated derivative du/dc
    
    return(0);
}

/*** R
myfun2(1.0, 2.0, 3.0, 0.5)
*/
/*-------------------------------------------------------------------------------------------------
=== Note ===
---------------------------------------------------------------------------------------------------
This example would also work if real was used instead of dual. Should you
need higher-order cross derivatives, however, e.g.,:

    double d2udxda = derivative(f, wrt(x, params.a), at(x, params));

then higher-order dual types are the right choice since real types are
optimally designed for higher-order directional derivatives.
-------------------------------------------------------------------------------------------------*/

```

## Calculating Gradient vector of a scalar function

Differentiation a scalar output from a vector input
$$
f(x) = \sum_{i = 1}^{5} (x_i \cdot e^{x_i})
$$
We also see use of `Eigen` vectors in the example below, so instead of 

```{Rcpp, eval = FALSE}
//[[Rcpp::depends(Rcppautodiff)]]
```
we write, 
```{Rcpp, eval = FALSE}
//[[Rcpp::depends(RcppEigen, Rcppautodiff)]]
```

We also need to manually convert `Rcpp`'s `NumericVector` to `ArrayXreal`
```{Rcpp, eval = FALSE}
#include <Rcppautodiff.h>

// autodiff include
#include <autodiff/forward/real.hpp>
#include <autodiff/forward/real/eigen.hpp>

using namespace Rcpp;
using namespace autodiff;

// [[Rcpp::plugins(cpp17)]]

//[[Rcpp::depends(RcppEigen, Rcppautodiff)]]

// The scalar function for which the gradient is needed
real f3(const ArrayXreal& x)
{
    return (x * x.exp()).sum(); // sum([xi * exp(xi) for i = 1:5])
}

// [[Rcpp::export]]
int myfun3(NumericVector x_input)
{
    using Eigen::VectorXd;

	// Convert input vector to ArrayXreal manually
    ArrayXreal x(x_input.length());             // declare the input array x
    for (int i = 0; i < x_input.length(); i++){
        x(i) = x_input[i];
    }

    real u;                                     // the output scalar u = f(x) evaluated 
                                                // together with gradient below

    VectorXd g = gradient(f3, wrt(x), at(x), u); // evaluate the function value u and its
                                                // gradient vector g = du/dx

    Rcpp::Rcout << "u = " << u << std::endl;    // print the evaluated output u
    Rcpp::Rcout << "g = \n" << g << std::endl;  // print the evaluated gradient vector g = du/dx
}

/*** R
myfun3(c(1,2,3,4,5))
*/
```

## Gradient vector of a scalar function with parameters

Differentiating the function
$$
f(x) = \sum_{i}^{N_1} x_i^2 + \sum_i^{N_2} p_{i} + q
$$
```{Rcpp, eval = FALSE}
#include <Rcppautodiff.h>

// autodiff include
#include <autodiff/forward/real.hpp>
#include <autodiff/forward/real/eigen.hpp>

using namespace Rcpp;
using namespace autodiff;

// [[Rcpp::plugins(cpp17)]]

//[[Rcpp::depends(RcppEigen, Rcppautodiff)]]

// The scalar function for which the gradient is needed
real f4(const ArrayXreal& x, const ArrayXreal& p, const real& q)
{
    return (x * x).sum() * p.sum() * exp(q); // sum([xi * xi for i = 1:5]) * sum(p) * exp(q)
}

// [[Rcpp::export]]
int myfun4(NumericVector x_input, NumericVector p_input, double q_input)
{
    using Eigen::VectorXd;

    ArrayXreal x(x_input.length());             // declare the input array x
    for (int i = 0; i < x_input.length(); i++){
        x(i) = x_input[i];
    }

    ArrayXreal p(p_input.length());             // declare the input array p
    for (int i = 0; i < p_input.length(); i++){
        p(i) = p_input[i];
    }

    real q = q_input;        // the input parameter q as a single variable

    real u;             // the output scalar u = f(x, p, q) evaluated together with gradient below

    VectorXd gx   = gradient(f4, wrt(x), at(x, p, q), u);       // evaluate the function value
                                                                // u and its gradient vector gx = du/dx
    VectorXd gp   = gradient(f4, wrt(p), at(x, p, q), u);       // evaluate the function value
                                                                // u and its gradient vector gp = du/dp
    VectorXd gq   = gradient(f4, wrt(q), at(x, p, q), u);       // evaluate the function value
                                                                // u and its gradient vector gq = du/dq
    VectorXd gqpx = gradient(f4, wrt(q, p, x), at(x, p, q), u); // evaluate the function value
                                                                // u and its gradient vector 
                                                                // gqpx = [du/dq, du/dp, du/dx]

    std::cout << "u = " << u << std::endl;          // print the evaluated output u
    std::cout << "gx = \n" << gx << std::endl;      // print the evaluated gradient vector gx = du/dx
    std::cout << "gp = \n" << gp << std::endl;      // print the evaluated gradient vector gp = du/dp
    std::cout << "gq = \n" << gq << std::endl;      // print the evaluated gradient vector gq = du/dq
    std::cout << "gqpx = \n" << gqpx << std::endl;  // print the evaluated gradient vector 
                                                    // gqpx = [du/dq, du/dp, du/dx]
}

/*-------------------------------------------------------------------------------------------------
=== Note ===
---------------------------------------------------------------------------------------------------
This example would also work if dual was used instead. However, if gradient,
Jacobian, and directional derivatives are all you need, then real types are your
best option. You want to use dual types when evaluating higher-order cross
derivatives, which is not supported for real types.
-------------------------------------------------------------------------------------------------*/
/*** R
myfun4(c(1,2,3,4,5), c(1,2,3), -2)
*/
```

